The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Tides or nature?

### 2025-09-09 thesis/results.tex
thesis updates

In this chapter, we present our results from our simulations of Sculptor
and Ursa Minor. In each case, we first discuss the range of reasonable
orbits and initial dark matter halos for each galaxy. Then, we describe
how tides affect the evolution of each galaxy. We furthermore consider
how the effects of the LMC changes the results, including using
additional N-body simulations in evolving potentials. Finally, we review
the predictions and characteristics of the present-day stellar
populations and properties from our models.

\section{Sculptor}\label{sculptor}

\subsection{Milky Way tides}\label{milky-way-tides}

\subsection{Tidal effects}\label{tidal-effects}

The Milky Way tides indeed affect Sculptor's dark matter halo.
Fig.~\ref{fig:scl_sim_images} shows projected dark matter distribution
of our N-body simulation for Sculptor in the \(y\)-\(z\) plane for 5
previous pericentres. This simulation uses the \texttt{compact} initial
halo on the \texttt{smallperi} orbit. The dark matter forms large
streams which orbit the Milky Way several times. However, note that the
central dark matter cusp is similar in appearance across all simulation
snapshots.

More quantitatively, the halo loses about 99\% of its initial dark
matter mass. This corresponds in a reduction from the initial
\(v_{\rm max} = 31\,\kms\) to \(v_{\rm max} = 22\,\kms\). However, the
stellar component of Sculptor is relatively unaffected.

@@

The analytically motivated break and tidal radii corroborate a weak
tidal effect on stars. In both cases, these radii work out to be
\(\gtrsim 100\) arcminutes, outside where we can measure the density
profile. In particular, because we have chosen the most extreme
observationally permissible orbit, it is unlikely, for similar
\(\Lambda\)CDM motivated initial conditions, that these radii would ever
approach where the break in Sculptor's density profile appears.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_images.png}}
\caption[Sculptor simulation snapshots]{Images of the dark matter
evolution over a selection of past apocentres and the present day
position. Limits range from -150 to 150 kpc in the \(y\)-\(z\)
(approximately orbital) plane and the colourscale is logarithmic
spanning 5 orders of magnitude between the maximum and minimum values.
In this image, stars occupy only ever a few pixels so are not
plotted.}\label{fig:scl_sim_images}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.pdf}}
\caption[Sculptor Tidal Tracks]{The tidal tracks for the smallperi
orbit. Todo: add velocity dispersion plot to
RHS}\label{fig:scl_tidal_track}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sigma_v_time.pdf}}
\caption[Sculptor velocity dispersion evolution]{Evolution of stellar
velocity dispersion within 1 kpc for different Scl models. In all cases,
the evolution is mild. Note that binarity may reduce the inflate the
observed velocity dispersion by \textasciitilde{} 1 km/s, so the
conservative lower limit is around 8 km/s. \textbf{TODO}, add bound mass
with time. Maybe combine with tidal tracks and radius / time orbit.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_smallperi_i_f.pdf}}
\caption[Sculptor initial and final density profiles]{Effects on
exponential initial stars. TODO: plot 2D sky proj.
stars}\label{fig:scl_smallperi_i_f}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_plummer_i_f.pdf}}
\caption[Sculptor Plummer initial and final density profiles]{effects on
Plummer initial stars.}\label{fig:scl_smallperi_plummer_i_f}
\end{figure}

@@

The Milky Way isn't the only galaxy in town. Becoming more apparent, the
infall of the LMC substantially affects the Milky Way system. The mass
of the LMC may be as high as 1/5 of the Milky Way mass, so including an
LMC in the Milky Way potential may change conclusions about properties
and orbits of the Milky Way satellite system.

@@

As discussed in \citet{battaglia+2022}, Sculptor's orbit is strongly
influenced by the presence of an LMC.

To explore the effects of the LMC on Sculptor, we use the publically
available L3M11 model from \citet{vasiliev2024}. This model stars with a
lighter Milky Way than our fiducial \citet{EP2020}, \citet{mcmillan2011}
like model. The L3M11 potential is an evolving multipole approximation
of an N-body simulation with both a live MW and LMC dark matter halo.
The MW intital conditions were an NFW with \(r_s=16.5\,\)kpc and mass
\(M_{\rm 100}= 11\times10^{11}\Mo\), and the LMC was a NFW halo with
\(r_s=11.7\) and \(M_{100} = 2.76 \times 10^{11} \Mo\). This model had a
previous LMC pericentre at about 6 Gyr ago.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_xyzr_orbits.png}}
\caption[Sculptor Orbits with LMC]{This figure is similar to
Fig.~\ref{fig:scl_orbits} except that we are showing the orbits with and
without an LMC. In the bottom row, the distance from Sculptor (or the
LMC) to the MW is plotted (left), and the Sculptor - LMC distance
(right.)}\label{fig:scl_lmc_orbits_effect}
\end{figure}

\textbf{tinyperilmc}

\begin{itemize}
\item
  ra = 15.0183 dec = -33.7186 distance = 73.1 pmra = 0.137 pmdec =
  -0.156 radial\_velocity = 111.2

  t\_i = -2.00 pericentre = 38.82 apocentre = 187.50 t last peri = -0.33
  x\_i = {[}4.30 138.89 125.88{]} v\_i = {[}6.89 -56.83 52.09{]}
\end{itemize}

\subsubsection{Tidal effects}\label{tidal-effects-1}

@@

A final note is that this model only considers the recent tidal effects
.As illustrated in the range of possible orbits in
Fig.~\ref{fig:scl_lmc_orbits_effect}, there is a chance that Sculptor
experienced an extremely small pericentre with the Milky Way. This
pericentre has the potential to substantially rearrange the stellar
component, drive large tidal mass loss, and create a Plummer-like
stellar density profile. However, this encounter is highly dependent on
the choice of the MW-LMC potential model, and may not occur at all. Long
time integration in dynamic potentials amplifies uncertainties in the
inputs and may not be reliable. \textbf{TODO: I would love to get this
model to work\ldots{}}

@@

>>>>

As discussed in Chapters \ref{sec:introduction}, \ref{sec:data},
Sculptor and Ursa Minor have unusually extended stellar light profiles.
Using our simulation setup introduced in Chapter \ref{sec:methods}, we
present our simulation results for each galaxy in this Chapter. We find
that tides may drive significant evolution of each galaxies total mass.
However, the compact stellar distribution for each galaxy is nearly
unaffected. We also show that the LMC strongly perturbs Scl's orbit.
But, the tidal effects when considering an LMC are yet weaker than in
the Milky Way only case. We conclude that tides are unlikely to affect
the stellar component of either galaxy.

\subsection{Tidal evolution in the Milky
Way}\label{tidal-evolution-in-the-milky-way}

The Milky Way tides drive significant mass loss for Sculptor.
Fig.~\ref{fig:scl_sim_images} shows projected dark matter distribution
of our N-body simulation for Sculptor in the \(y\)-\(z\) plane for 5
previous pericentres. This simulation uses the fiducial initial halo on
the \smallperi{} orbit. The dark matter forms large streams which orbit
the Milky Way several times. Note that, in contrast to the outer
regions, the central dark matter cusp is similar in appearance across
all simulation snapshots.

Fig.~\ref{fig:scl_tidal_track} shows the initial and final circular
velocity profile of the dark matter, and the evolution of the maximum
circular velocity. Sculptor's final dark matter halo evolves from
\(\vmax=31\,\kms\) at \(\rmax=3.2\,\kms\) to \(\vmax = 22\,\kms\) at
\(\rmax = 2\,\kpc\). In terms of bound mass, the mass evolves from
\(0.40 \times 10^{10}\,\Mo\) to \(0.026\times10^{10}\,\Mo\). The
circular velocity maximum evolves along the tidal track predicted by
\citet{EN2021}.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_images.png}}
\caption[Sculptor simulation snapshots]{Images of the dark matter
evolution over a selection of past apocentres and the present day
position. Limits range from -150 to 150 kpc in the \(y\)-\(z\)
(approximately orbital) plane and the colourscale is logarithmic
spanning 5 orders of magnitude between the maximum and minimum values.
The green dot represents the final (observed) position of the galaxy and
the grey curve represents the orbit over one radial
oscillation.}\label{fig:scl_sim_images}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.png}}
\caption[Sculptor Tidal Tracks]{The tidal tracks for the smallperi
orbit. Todo: add velocity dispersion plot to RHS, label initial and
final with times}\label{fig:scl_tidal_track}
\end{figure}

\subsection{}\label{section}

@@

The analytically motivated break and tidal radii also support a weak
tidal effect on stars. In both cases, these radii work out to be
\(\gtrsim 100\) arcminutes, outside where we can measure the density
profile. In particular, because we have chosen the most extreme
observationally permissible orbit, it is unlikely, for similar
\(\Lambda\)CDM motivated initial conditions, that these radii would ever
approach where the break in Sculptor's density profile appears.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_smallperi_i_f.pdf}}
\caption[Sculptor initial and final density profiles]{The tidal effects
on Scl's stellar component, for the \smallperi{} orbit with the fiducial
halo and exponential stars with \(R_s=0.10\,\kpc\). \textbf{Left:} the
final 2D projected density of stars on the sky, with colours
representing logarithmically increasing density. The circle marks the
break radius. \textbf{Right:} The initial (dotted) and final (solid)
stellar density profiles as compared to the observed stellar density
profile. Arrows mark the break and Jacobi (Eq.~\ref{eq:jacobi_radius})
radii. \textbf{add initial sky-projected density, t=xxx in legend,
sculptor in figure title}}\label{fig:scl_smallperi_i_f}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_plummer_i_f.pdf}}
\caption[Sculptor Plummer initial and final density profiles]{Similar to
Fig.~\ref{fig:scl_smallperi_i_f} except for Plummer initial stars with
\(R_h = 0.20\,\kpc\). While a faint stream may be visible with deeper
observations, effects on the stellar profile are minimal. \emph{could
overlay orbit on left}}\label{fig:scl_smallperi_plummer_i_f}
\end{figure}

@@

The Milky Way isn't the only galaxy in town. Recently, work has shown
that the infall of the LMC may substantially affects the Milky Way
system
\citep[e.g.,][]{erkal+2019, cautun+2019, garavito-camargo+2021, vasiliev2023}.
The mass of the LMC may be as high as 1/5 of the Milky Way mass.
Including an LMC in the Milky Way potential may change conclusions about
properties and orbits of the Milky Way satellite system, including Scl
\citep[e.g.,][]{patel+2020, battaglia+2022}.

@@

To explore the effects of the LMC on Sculptor, we use the publicly
available L3M11 model from \citet{vasiliev2024}. While
\citet{vasiliev2024} present other models, the L3M11 model has the
highest MW and LMC masses, and the (recent) orbital history appears to
be independent of the potential details. This model stars with a lighter
Milky Way than our fiducial, \citet{mcmillan2011-like} model. The L3M11
potential is an evolving multipole approximation of an N-body simulation
with both a live MW and LMC dark matter halo. The potential additionally
includes the reflex motion of the MW due to the LMC. The MW was
initialized as an NFW dark matter halo with \(r_s=16.5\,\)kpc and mass
\(M_{\rm 100}= 11\times10^{11}\Mo\), and the LMC was a NFW halo with
\(r_s=11.7\) and \(M_{100} = 2.76 \times 10^{11} \Mo\). This model had a
previous LMC pericentre at about 6 Gyr ago.

For our orbit, we select the orbit with the \(3\sigma\)-smallest LMC
pericentre. Results are similar when selecting for a small MW pericentre
instead. We also use, instead, the \texttt{small} halo for Sculptor, as
the tidal effects do not reduce the velocity dispersion, unlike for the
MW-only model.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_xyzr_orbits.png}}
\caption[Sculptor Orbits with LMC]{This figure is similar to
Fig.~\ref{fig:scl_orbits} except that we are showing the orbits with and
without an LMC. In the bottom row, the distance from Sculptor (or the
LMC) to the MW is plotted (left), and the Sculptor - LMC distance
(right.) \textbf{cut scale on buttom
panel}}\label{fig:scl_lmc_orbits_effect}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_orbits_mass_effect.png}}
\caption[Sculptor Orbits with LMC]{The long-term orbital history of
Sculptor is uncertain. In light, transparant lines, the orbits of
Sculptor in three different LMC/MW mass models from \citet{vasiliev2024}
are shown for random samples of the observed properties. The LMC orbits
are in solid, thick lines of the corresponding colour. The L2M11 has a
lighter LMC mass, and the L3M10 model has a lighter MW mass than our
fiducial L3M11 LMC model.}\label{fig:scl_lmc_orbits_mass}
\end{figure}

\begin{table*}[t]
\centering
\caption[Sculptor Selected Orbits]{Properties of selected orbits for Sculptor. The mean orbit represents the observational mean from Table \ref{tbl:scl_obs_props}. The \smallperi{} represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects. }
\label{tbl:scl_orbits}
\begin{tabular}{ll}
\toprule
Property & smallperi-LMC\\
\midrule
distance / kpc & 73.1\\
$\pmra / \masyr$ & 0.137\\
$\pmdec / \masyr$ & –0.156\\
LOS velocity / $\kms$ & 111.2\\
$t_i / \Gyr$ & -2\\
${x}_{i} / \kpc$ & 4.30\\
${y}_{i} / \kpc$ & 138.89\\
${z}_{i} / \kpc$ & 125.88\\
$\V_{x\,i} / \kms$ & 6.89\\
$\V_{y\,i} / \kms$ & -56.83\\
$\V_{z\,i} / \kms$ & 52.09\\
pericentre / kpc & 38.82\\
peri-LMC / kpc & ?\\
apocentre / kpc & 187.50\\
$t_{\rm last\ peri} / {\rm Gyr}$ & -0.46\\
$t_{\rm last\ peri-LMC} / \Gyr$ & -0.1?\\
Number of pericentres & 1-2\\
\bottomrule
\end{tabular}
\end{table*}

\subsubsection{Tidal effects}\label{tidal-effects}

@@

@@

\subsubsection{A possible second MW
pericentre}\label{a-possible-second-mw-pericentre}

A final note is that this model only considers the recent tidal effects
.As illustrated in the range of possible orbits in
Fig.~\ref{fig:scl_lmc_orbits_effect}, there is a chance that Sculptor
experienced an extremely small pericentre with the Milky Way. This
pericentre has the potential to substantially rearrange the stellar
component, drive large tidal mass loss, and create a Plummer-like
stellar density profile. However, this encounter is highly dependent on
the choice of the MW-LMC potential model, and may not occur at all. Long
time integration in dynamic potentials amplifies uncertainties in the
inputs and may not be reliable.

To illustrate the possible effects of this previous pericentre, we
create a model with a goal of matching the \(3\sigma\) smallest
pericentre with the MW. Because of strong tidal shocking, the final
conditions depend strongly on the initial conditions. In addition,
actions are not conserved in this evolving potential, so we use the
\citet{vasiliev2024} method of iteratively updating the initial
conditions by empirically estimating the Jacobian. Our final position
reaches adequate, but not perfect, agreement with the intended final
position.

As a result of this proceedure, while the point particle reaches a
pericentre of \(\sim 4\,\kpc\), the adjusted N-body orbit instead only
reaches a pericentre of \(\sim 12\,\kpc\).
